---
title: "Mapping Toronto's Libraries"
title-block-banner: true
subtitle: "Are Libraries Distributed Equally Throughout the City?"
author: Hadi Ahmad
thanks: "Code and data are available at: https://github.com/hadi-q/toronto-public-libraries."
date: today
date-format: "D MMMM YYYY"
abstract: "This report analyzes the Library Branch General Information dataset, which is made available to the public from Open Data Toronto. It uses the library branches, square footage, and ward information to determine how well libraries are spread across the city. In summary, (TO DO - KEY FINDINGS). These findings are relevant because they can help ascertain how accessible libraries are to residents of Toronto, and guide policy decisions on where new locations should be opened."

format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# force graph labels to standard notation
# copied from https://stackoverflow.com/questions/14563989/force-r-to-stop-plotting-abbreviated-axis-labels-scientific-notation-e-g-1e
options(scipen=10000)

# install.packages("tidyverse")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("readr")
# install.packages("janitor")
# install.packages("opendatatoronto")
# install.packages("knitr")
# install.packages("here")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(janitor)
library(opendatatoronto)
library(knitr)
library(here)

#### Read in cleaned data ####
analysis_data = read_csv(
  file = here("outputs/data/analysis_data.csv"),
  show_col_types = FALSE
)
```

# Introduction

Libraries are a critical form of social infrastructure in modern society. They serve many functions, ranging from providing access to books and WiFi, to programs on self-improvement, to providing places of shelter to study or work. More importantly, however, they offer most or all of this functionality for free. Thus, libraries are not just places for any specific demographic: they act as third spaces, allowing users of different ages, ethnicities, and backgrounds to all exist in the same space. In a society where most services are monetized or paid, libraries remain a vestige of a more egalitarian era. 

The Toronto Public Library in particular excels in these functions due to its size, cross-branch integration, and volume of services available. With 100 physical branches in its network, users can borrow books from any branch and have it delivered to their home branch. This increases the access to information that Toronto Public Library users benefit from. As well, users can borrow movies, internet hotspot devices, video game consoles, and even get free passes for museums or other attractions. A Toronto Public Library card also gives users access to third party platforms like Libby, OverDrive, Hoopla, and Consumer Reports, providing additional online resources that users can benefit from.

In October 2023, the Toronto Public Library was victim to a major cybersecurity attack that took down their website and online systems. This incident also led to some user data being compromised to hackers. As of January 2024, their website services remain offline. This prevents users from being able to borrow or renew books online, create or renew a library card, or access many of Toronto Public Library's services. As a result, physical proximity to a library branch has become more important than ever before, as it remains the only way to borrow books or access other library services until full services are restored.

This paper aims to analyze the distribution of Toronto Public Library branches across the 25 wards of Toronto. The distribution of branches will be viewed in two ways: number of branches and square feet of branch space within a given ward. The resulting analyses will determine if certain wards are over- or underrepresented in terms of the quantity and size of library branches, which could provide cursory information on where future libraries could be built or current libraries that could benefit from expansion. Future analyses may explore demographic factors to determine if libraries are disproportionately located in neighbourhoods of higher or lower income, to get a sense of equality of access to libraries.

The Data @sec-data of this paper explores the `opendatatoronto` data holistically, outlining tools and methods used to analyze the data and some preliminary observations from the data. The Results @sec-results introduces more observations found from the data analysis. The Discussion @sec-disc ties back the findings to its real world relevance. Lastly, the Conclusion @sec-concl summarizes key findings and outlines future areas of study.

# Data {#sec-data}

All relevant data was sourced from the Open Data Toronto portal, and extracted using the `opendatatoronto` library for `R` [@rOpendatatoronto]. In particular, one dataset was used for the graphs and analyses in this paper, which included information about all Toronto Public Library branches, their addresses, phone numbers, square footage, year of construction of the building, among other variables like whether there as a park adjacent to the library, if adult literacy classes are available, and GPS coordinates of the building.

## Data Tools

The data was generated, extracted, and cleaned using `R` [@citeR], leveraging functions from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `janitor` [@rJanitor], `opendatatoronto` [@rOpendatatoronto], `knitr` [@rKnitr], and `here` [@rHere].

## Overview of Dataset

The selected dataset was published by the Toronto Public Library to share information on libraries located in its network, and was last updated on 29 June, 2023 in accordance with its annual updating guideline. The chosen variables of analysis were the branch name, the square footage of the library,	the ward number and	ward name of where the library is located, and the year of construction of the site. See `Table 1` for an overview of the data.

```{r}
#| label: library-branches
#| fig-cap: "Table 1: Sample of Toronto Public Library Data"
#| echo: false

# Col names code copied from https://bookdown.org/yihui/rmarkdown-cookbook/kable.html
knitr::kable(head(analysis_data[, 1:5]), "simple",   col.names = c('Branch', 'Square Footage', 'Ward Number', 'Ward Name', 'Year Built'))

```

Libraries are present in each of the 25 wards in the city of Toronto. (Note that at the time of the 2018 election, wards were redrawn from 44 wards down to the present 25.) While neighbourhood data is available from the dataset, it is excluded from this analysis, as neighbourhood classifications and counts varied between different `opendatatoronto` datasets, and because not all neighbourhoods are listed in this dataset.

The original dataset had entries for codes that did not correspond to physical locations, such as the phone line. These were filtered out, leaving a total of 100 library branches.

## Distribution of Branches Per Ward by Count

On average, there were four library branches per ward. However, in practice, this ranges from one branch in wards like Willowdale, up to seven in branches like Toronto-Danforth. See `Table 2` for a sample of the dataset.

```{r}
#| label: library-branches-count
#| fig-cap: "Table 2: Sample of Count of Library Branches by Ward"
#| echo: false
#| 
# output how many libraries per ward
knitr::kable(head(analysis_data |>
  count(ward_name), 10), "simple",   col.names = c('Ward Name', 'Number of Branches'))

```
## Distribution of Branches Per Ward by Square Footage

There is significant range in the square footage of libraries. The smallest library in the dataset, Todmorden Room, measures just 554 square feet, while the largest, the Toronto Reference Library, amounts to over 400,000 square feet. Clearly, both of these branches are outliers from the dataset - the mean square footage of a library branch is 18,129.32 square feet, whereas the median is 8,496.50 square feet. This highlights the need to analyze the distribution of library branches beyond absolute counts, to better understand if library space is evenly allocated across the city. See `Table 3` for a sample of the dataset.

```{r}
#| label: library-branches-sq-ft
#| fig-cap: "Table 3: Sample of Total Square Footage of Library Branches by Ward"
#| echo: false

# output sq ft of libraries per ward
knitr::kable(head(analysis_data %>%
  group_by(ward_name) %>%
  summarize(sum_sq_ft = sum(square_footage, na.rm = FALSE)), 10), "simple",   col.names = c('Ward Name', 'Total Square Footage of Branches'))

```

## Merged Branch Count and Square Footage

To better understand any links between the number of branches and square footage of library space, the two datasets were merged into a single dataframe using the `merge` function.

```{r}
#| include: false
#| warning: false
#| message: false

# new dataframe with ward name and library count
analysis_data_by_ward_count <-
  analysis_data |>
    count(ward_name)

# new dataframe with ward name and total sq ft
# code copied from https://www.statology.org/sum-by-group-in-r/
analysis_data_by_ward_sq_ft <-
  analysis_data %>%
  group_by(ward_name) %>%
  summarize(Freq = sum(square_footage))

# merge dataframes
# code copied from https://stackoverflow.com/questions/1299871/how-to-join-merge-data-frames-inner-outer-left-right
analysis_data_by_ward <-
  merge(analysis_data_by_ward_count, analysis_data_by_ward_sq_ft, by = "ward_name")

```

```{r}
#| label: library-branches-by-ward-count-and-sq-ft
#| fig-cap: "Table 4: Count and Total Square Footage of Library Branches by Ward"
#| echo: false
# output merged dataframe
knitr::kable(
  analysis_data_by_ward,
  col.names = c('Ward Name', 'Number of Branches', 'Total Square Footage of Branches')
)

```


# Results {#sec-results}

The distribution of the number of branches per ward range from one branch, in Willowdale, to seven branches, in Toronto-Danforth. Overall, the distribution follows a normal curve, with 10 of the 25 wards falling squarely on the median and mode of having four branches. Only one ward each falls in either upper or lower extremity. See Figure 1 for more details.

```{r}
#| label: library-branches-by-ward-count-bar
#| fig-cap: "Library Branches by Ward by Count"
#| echo: false
# horizontal bar chart
# copied from https://stackoverflow.com/questions/10941225/horizontal-barplot-in-ggplot2
ggplot(analysis_data_by_ward_count, aes(x=reorder(ward_name, n), y=n)) +
  geom_bar(stat='identity', fill="lightblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Wards", y = "Number of Branches", color = "Country") +
  scale_color_brewer(palette = "Set1")

```

```{r}
#| label: library-branches-by-ward-sq-ft-bar
#| fig-cap: "Library Branches by Ward by Square Footage"
#| echo: false
# horizontal bar chart
# copied from https://stackoverflow.com/questions/10941225/horizontal-barplot-in-ggplot2
ggplot(analysis_data_by_ward_sq_ft, aes(x=reorder(ward_name, Freq), y=Freq)) +
  geom_bar(stat='identity', fill="lightblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Wards", y = "Total Square Footage", color = "Country") +
  scale_color_brewer(palette = "Set1")

```

# Discussion {#sec-disc}



# Conclusion {#sec-concl}



# LLMs

Statement on LLM usage: no LLMs were used in the making of this paper.

# References
